SSA based global code motion of identical computations

### 2017-04-17 paper.tex
fix section 3

The algorithm for code motion uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Dominance (DOM) and Post-Dominance (PDOM) relations \cite{dragonbook} on a
  Control Flow Graph (\CFG{}).
\item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
  dominator tree with join-edges to keep track of data-flow in a program. We use
  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012}.
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given
  a unique number and the expressions that the compiler can prove to be
  identical are given the same number;
\item Memory \SSA{} \cite{novillo2007memory}: it is a factored use-def chain of
  memory operations that the compiler is able to prove are dependent.
\end{itemize}

@@

\subsection{Liveness using DJ Graph}
DJ-Graph \cite{Sreedhar1996} is a data structure that augments the dominator
tree with join-edges to keep track of data-flow in a program. We use DJ-Graph
and Merge Sets to compute liveness of variables as illustrated in
\cite{das2012}.  It is very efficient for computing liveness and does not
require any bitvectors to be maintained for each basic block. The underlying
simplicity of liveness computation is due to \SSA{} form, where the the values only
flow (from def to use) either through dominator edges or the join edges (where
we insert a PHI). The DJ-graph contains both these edges which allows for
computation of merge sets for each basic block i.e., a set of all basic blocks
where the values can flow from a particular basic block. We have implemented
merge-set computation from DJ-graph as illustrated in \cite{das2005}. A simplified
version of merge set that we implemented is presented here:

\begin{verbatim}
// Return true if the merge set of source node of
// a visited J-edge (incoming edge of lnode) is not
// the subset of the merge set of lnode.
bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet M)
   for (all incoming edges to lnode) do
     Let e = Incoming edge
     if (e is in JE && e in V) then
       Let snode = Source Node of e
       if (M(snode) is not a subset of M(lnode)) then
         return true
       end if
     end if
   end for
return false

// Compute merge set top-down in breadth first order.
bool constructMergeSet_1(BFSList B, JEdges JE, DomLevel DL)
repeat = False
// List of visited edges
Visited V
for n in B do
  for e in (all incoming edges to n) do
    if (e is in JE && e not in V) then
      V(e) = true
      Let snode = Source Node of e
      Let tnode = Target Node of e
      Let tmp = snode
      Let lnode = NULL
      while (DL(tmp) ≥ DL(tnode)) do
        Merge(tmp) = Merge(tmp) U Merge(tnode) U {tnode}
        lnode = tmp
        tmp = dom-parent(tmp)
      end while
      repeat = still_inconsistent(lnode, JE, Merge)
    end if
  end for
end while
return repeat

// Construct Merge set of each node in control-flow-graph G of a function.
void  constructMergeSet(CFG G) {
  B = Breadth First Order of G
  JE = JEdges of G
  DL = List of Path length (from root) of each node in G.

  do // Call until a fixed point is reached.
    Repeat = constructMergeSet_1(B, JE, DL);
  while (Repeat);
}
\end{verbatim}

We compute merge sets of the control flow graph which remains same through the
code-motion transformation. For code-motion we only want to know if a use
operand is a kill (to compute changes in register pressure). For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoistable/sinkable
candidates is to be evaluated. A simplified version of isLiveOutUsingMergeSet is
presented here:

\begin{verbatim}
bool isLiveOutUsingMergeSet()
  // Compute if variable a is liveout from basic block n
  Input: Node N, Variable a

  if a is defined in N then
    if a is used outside any basic block other than N then
      return true;
    else return false;
  endif

  Ms(n) = null;
  // Mergeset of N is the union of merge sets of its successors
  for w in successors(n) do
    Ms(n) = Ms(n) U Mr(w);
  endfor

  // Iterate over all the uses of a and see if any intersect with the
  // merge set of N.
  for t in users(a) do
    b = basic_block(t)
    while (b != null) and (b != def_bb(a)) do
      if b ∩ Ms(n) then
        return true;
      endif
      b = dom-parent(b);
    endwhile
  endfor
  return false;
\end{verbatim}

\begin{verbatim}
// return true if op is the last use of the
// operand in I.
bool isKill(Operand op, Instruction I) {
  if (isLiveOutUsingMergeSet(op))
    return false;
  B = basic_block(I);
  for each I1 after I in B:
    if (I1 uses op)
      return false;

return true;
}
\end{verbatim}

@@

Compute GVN of each expression in the function
Compute DJ Graph of function and compute mergeset based on that

@@

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that are
candidates for code-motion, matching the complexity of \PRE{} on \SSA{} form.
The analyses computed for this pass are Global Value Numbering, Computation of
DJ Graph and MergeSets, Marking Barriers, all are linear in number of
instructions in a function. Liveness analysis expensive but only performed
on-demand for hoistable candidates so it does not affect the overall compile
time by much. Other analyses like Alias Analysis, \MemorySSA{} and Dominator Tree
are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
Barriers for each iteration of the code-motion, we have not seen signinficant
increase in the compilation times. We have also provided appropriate compiler
flags to expedite code-motion by bailing out with fewer iterations, or skip the
liveness based profitability analysis to aggressively move the code as long as
they are legal (Section~\ref{sec:experimental-results}).

>>>>

The algorithm for code motion uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Dominance (DOM) and Post-Dominance (PDOM) relations \cite{dragonbook} on a
  Control Flow Graph (\CFG{});
\item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
  dominator tree with join-edges to keep track of data-flow in a program. We use
  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012};
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given
  a unique number and the expressions that the compiler can prove to be
  identical are given the same number;
\item \MemorySSA{} \cite{novillo2007memory}: it is a factored use-def chain of
  memory operations that the compiler is unable to prove
  independent. \MemorySSA{} accelerates the access to the alias analysis
  information.
\end{itemize}

@@

\subsection{Liveness using DJ-graph}
DJ-graph \cite{Sreedhar1996} is a data structure that augments the dominator
tree with join-edges to keep track of data-flow in a program. We use DJ-graph
and merge-sets to compute liveness of variables as illustrated in
\cite{das2012}.  It is very efficient for computing liveness and does not
require any bitvectors to be maintained for each basic block. The underlying
simplicity of liveness computation is due to \SSA{} form, where the the values
only flow (from def to use) either through dominator edges or the join edges
(where we insert a PHI). The DJ-graph contains both these edges which allows for
computation of merge-sets for each basic block, i.e., a set of all basic blocks
where the values can flow from a particular basic block. We have implemented the
merge-set computation based on a DJ-graph as illustrated in \cite{das2005}. A
simplified version of the merge-set that we implemented is presented here:

\begin{verbatim}

// Return true if the merge-set of source node of a visited J-edge (incoming edge of
// lnode) is not the subset of the merge-set of lnode.
bool still_inconsistent(Node lnode, JEdges JE, Visited V, MergeSet Merge)
  for e in (all incoming edges to lnode)
    if (e is in JE && e in V)
      if (Merge(Source Node of e) is not a subset of Merge(lnode))
        return true
  return false

// Compute a merge-set top-down in breadth first order.
bool constructMergeSet_1(BFSList B, JEdges JE, DomLevel DL)
  repeat = false
  V = empty list of visited edges
  for n in B
    for e in (all incoming edges to n)
      if (e is in JE && e not in V)
        V(e) = true
        let tmp = Source Node of e
        let tnode = Target Node of e
        let lnode = NULL
        while (DL(tmp) >= DL(tnode))
          Merge(tmp) = Merge(tmp) U Merge(tnode) U {tnode}
          lnode = tmp
          tmp = dom-parent(tmp)
        repeat = still_inconsistent(lnode, JE, Merge)
  return repeat

// Construct merge-set of each node in Control-Flow-Graph G of a function.
void  constructMergeSet(CFG G) {
  B = Breadth First Order of G
  JE = JEdges of G
  DL = List of Path length (from root) of each node in G.

  do // Call until a fixed point is reached.
    Repeat = constructMergeSet_1(B, JE, DL)
  while (Repeat)
}
\end{verbatim}

We compute merge sets of the control flow graph which remains the same through
the code-motion transformation. For code-motion we only want to know if a use
operand is a kill (to compute changes in register pressure.) For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoistable/sinkable
candidates is to be evaluated. A simplified version of isLiveOutUsingMergeSet is
presented here:

\begin{verbatim}
// Return true if variable A is liveout from basic block N
bool isLiveOutUsingMergeSet(Node N, Variable A)
  if (A is defined in N)
    if (A is used outside any basic block other than N)
      return true
    return false

  Ms(N) = null
  // Mergeset of N is the union of merge sets of its successors
  for W in successors(N)
    Ms(N) = Ms(N) U Mr(W)

  // Iterate over all the uses of A and see if any intersect with the merge-set of N.
  for T in users(A)
    B = basic_block(T)
    while (B != null and B != def_bb(A))
      if B in Ms(N) return true
      B = dom-parent(B)
  return false

// Return true if op is the last use of the operand in I.
bool isKill(Operand op, Instruction I)
  if (isLiveOutUsingMergeSet(op))
    return false

  for each I1 after I in basic_block(I)
    if (I1 uses op)
      return false
  return true
\end{verbatim}

@@

Compute GVN of each expression in the function
Compute DJ-graph of function and compute mergeset based on that

@@

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that are
candidates for code-motion, matching the complexity of \PRE{} on \SSA{} form.
The analyses computed for this pass are Global Value Numbering, Computation of
DJ-graph and merge-sets, Marking Barriers, all are linear in number of
instructions in a function. Liveness analysis expensive but only performed
on-demand for hoistable candidates so it does not affect the overall compile
time by much. Other analyses like Alias Analysis, \MemorySSA{} and Dominator Tree
are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
Barriers for each iteration of the code-motion, we have not seen signinficant
increase in the compilation times. We have also provided appropriate compiler
flags to expedite code-motion by bailing out with fewer iterations, or skip the
liveness based profitability analysis to aggressively move the code as long as
they are legal (Section~\ref{sec:experimental-results}).
